Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-23 response-to-reviewers.tex

% -*- mode: LaTeX; mode: flyspell; -*-
% mode: Tex-Pdf -*-
\documentclass{article}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[bookmarks=TRUE,
            colorlinks,
            pdfpagemode=none,
            pdfstartview=FitH,
            citecolor=black,
            filecolor=black,
            linkcolor=blue,
            urlcolor=black,
            ]{hyperref}
\usepackage[top=15mm, bottom=25mm, left=10mm, textwidth=130mm,
marginparwidth=70mm]{geometry}
\usepackage{graphicx}
\usepackage{icomma}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{setspace}
\usepackage[dvipsnames]{xcolor}
\usepackage{xspace}
\usepackage{csquotes}
\usepackage[layout=margin,author=,status=draft]{fixme}

@@

\fxsetup{theme=color}
\newcommand{\skb}[1]{
	\begin{quotation}
		\color{Blue} SKB: #1	
	\end{quotation}
}
\newcommand{\debug}[1]{\marginpar{\small\raggedright\begin{spacing}{1}
      \color{BrickRed} #1
\end{spacing}
}}
\newenvironment{response}
{\slshape}{}

\begin{document}
\section*{Response to the referees}

\section{Comments by the editor}

\begin{enumerate}
\item In addition, please elaborate further the comparison of your
  results with respect to the previous literature about the role of
  language in Baltic, CEE, and former Soviet Union countries.

  \begin{response}
    We have added some more references about CEE economies.  This
    literature is now discussed in introduction and our outcomes are
    benchmarked against some of the CEE literature in the discussion. 
  \end{response}
\item 
In revising your paper, please make an effort to keep within the
journal's word-length norm of 7500 words (including references, notes
and appendices, but not figures and tables) and restrict the number of
references to 50 items.

\debug{nothing else but 7500 words/50 items.}
\item When submitting your revised manuscript, you will be able to respond to the comments made by the referee(s) in the space provided.  You can use this space to document any changes you make to the original manuscript.  In order to expedite the processing of the revised manuscript, please be as specific as possible in your response to the referee(s).

@@

\section{Referee 1}

\subsection{Main comments:}

\begin{enumerate}
\item The paper aims to contribute to an already large
  literature. The recent literature in this field focus on the
  identification of causal effects by using instrumental variable
  approaches (see, e.g., Isphording et al., 2014; Yao and van Ours,
  2015). However, I like the fact that the authors a very clear that
  they are not able to identify causal effects. This is, of course, a
  shortcoming and is limiting the overall contribution of the
  paper. Given the specific focus of International Journal of Manpower
  and the interests of its readership, I find that the paper is
  nevertheless providing a small, but still interesting
  contribution. By analysing data from Estonia and providing results
  on natives’ language skills the paper distinguished itself form
  large parts of the literature and adds some new results to our
  knowledge on the labour market effects of language skills.

  \begin{response}
    No response needed here.
  \end{response}

\item 
  My main concern is with some of their wage results. I find it
  puzzling that they don’t find a wage premium for Russian men who
  have Estonian language skills as they do find a sizeable significant
  effect of about 11 to 17 percent for Russian women and a significant
  positive effect for Estonians who speak Russian. Furthermore, they
  find that Estonian language skills do reduce the probability of
  being unemployed for Russians. Why is there a reduction in the
  probability of being unemployed, but no wage effect? Can we trust
  this result? Looking at the descriptive statistics (see Table 1) the
  wage for Russian men who speak Estonian seems to be higher compared
  to Russian men without the ability to speak Estonian (489.96 vs
  401.59). The authors should either provide more statistics on this
  result (if possible) or a more detailed discussion on potential
  reasons that lead to this finding.

  \begin{response}
    This is a good point.  We have added some more analysis (see
    Section 7), and it appears the story is somewhat more nuanced.
  \end{response}
\item The paper needs to be checked for spelling and grammar
  errors. There are a number of typos and errors, which should be
  corrected in a revision. This would normally be a minor comment, but
  here the authors should take this comment very serious as there are
  so many mistakes.

  \begin{response}
    We have done a thorough proofreading and improved the language. 
  \end{response}
\end{enumerate}

### 2018-07-23 paper/ijm.tex
First round of tool assisted spell checking; Renaming of files due to journal requirements

\fxsetup{theme=color}
%Custom commands
\newcommand{\eur}{\EUR}
\newcommand{\std}[1]{\emph{(#1)}}
\newcommand{\V}{{\ensuremath\checkmark}}
\def\fns{\footnotesize}
\renewcommand*{\vec}[1]{\boldsymbol{#1}}

@@

%\maketitle
%\begin{center}
%	\textbf{{\large Language Skills in an Ethnically Segmented Labour market:
%				Estonia 1989 -- 2012}} \\
%	{\large Sven-Kristjan Bormann} \\
%	\emph{School of Economics and Business Administration, University of Tartu, Narva mnt. 4, 51007 Tartu, Estonia\\ Tel.: +372 5844 3873\\ Email: \href{mailto:bormann@ut.ee}{bormann@ut.ee}}	\\
%
%	{\large Svetlana Ridala} \\
%	\emph{School of Business and Governance, Tallinn University of Technology,  Akadeemia tee 3, 12618 Tallinn, Estonia\\ Email: \href{
%			mailto:svetlana.ridala@gmail.com}{
%			svetlana.ridala@gmail.com}} \\
%		{\large Ott Toomet}\\
%		\emph{Information School, University of Washington, USA; Email: \href{mailto:otoomet@uw.edu}{otoomet@uw.edu}}
%%	\emph{(Received XX XXXX 20XX; accepted XX XXXX 20XX)}\\
%	\textbf{Corresponding author:} Sven-Kristjan Bormann Email: \href{mailto:bormann@ut.ee}{bormann@ut.ee}
%\end{center}
%\paragraph{Short biographical note:} \hspace{0pt} \\
%Sven-Kristjan Bormann born 28.11.1985, graduated in 2011 from Kiel University, Germany in Economics with specialisation in Quantitative Economics and Computer Science. Since September 2012 PhD-student in Economics at Tartu University, Estonia. \\
%Svetlana Ridala born 28.03.1976, graduated in 2003 from Tallinn Pedagogical University, Estonia master's (scien) degree in mathematics. Since 2009 PhD-student in Economics at Tallinn School of Economics and Business Administration, Estonia.
%\\
%Ott-Siim Toomet born 18.03.1970, graduated in 2004 from Aarhus University, Denmark PhD in Economics. Since 2015 Visiting scholar, Information School, University of Washington.
%
%\paragraph{Research funding}
%This research was supported by European Social Fund's Doctoral Studies
%and Internationalisation Program DoRa through Archimedes Foundation, and the Estonian Science Foundation grant 8627.
%\newpage
%\noindent
{\centering \textbf{Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012}}
\begin{abstract}
  \noindent
  \textbf{Purpose --} We analyze the relationship between skills of
  Estonian, Russian and English, and labour market
  outcomes in Estonia, a linguistically divided country.
  \\
  \textbf{Design/methodology/approach --} We use the Estonian Labour
  Force Surveys 1992--2012.  We rely on multivariate linear regression
  models to document the relationship between language skills and
  labor market outcomes.
  \\
  \textbf{Findings --} Estonian language knowledge (for ethnic Russians)
  are important determinants of unemployment.  Wage, in contrary, is
  closely related to English skills.  Ethic Russian men do not earn
  any premium from speaking Estonian, while women, fluent in Estonian
  earn approximately 10\% more.  For ethnic Estonians, Russian fluency
  is associated with a similar income gain.
  \\
  \textbf{Research limitations/implications -- } Due to the
  observational nature of the data, the effects reported in our study
  are not causal effects.  As a second limitation, the self-reported
  language skills data may be imprecise and hence the effects we
  report may be too small.
  \\
  \textbf{Practical implications --} Our results stress the role of
  workplace segregation, both along gender and ethnic lines, in
  determining the individual labor market experience.  
  \\
  \textbf{Originality/value --} We provide a comprehensive overview of
  the effects of language skills in a rapidly developing labor
  market in a linguistically divided economy.  We analyze several
  languages with different legal status and document long-term
  trends in the effects.
  \\
\end{abstract}
\textbf{Keywords:} language skills, ethnicity, wage gap, unemployment, Estonia \\
\textbf{JEL classification:} J15, J31, J71

@@

There is also an increasing body of literature focusing on other languages than the local main business language.  These studies generally find
that English is a significant predictor of better income
\citep{Lang2009, Casale2011, Toomet2011, Williams2011, azam+2013EDandCC, isphording2013, fabo+2017E}.
The role of other languages, in particular in multi-lingual economy, is
less clear.  For instance,
\cite{Drinkwater1997} finds
a positive effect for Welsh speaking workers in
Wales and \cite{Armstrong2015} shows that bilingual native French-speakers enjoy
a significant premium both inside and outside of Quebec.
\citet{FrancoisGrin1998}, in contrast,
shows that Italian language skills in Swiss non-Italian speaking
regions are associated with lower wage.  It remains unclear if
this is a true effect or just an omitted variable bias.

Few previous studies have provided a comprehensive analysis of a labour
market from the viewpoint of language skills.
\citet{Chiswick1995, chiswick+miller2002, chiswick+miller2007,
  Bellante1998, Chiswick2010} analyse the effects of dominant language
fluency for earnings among immigrants in the United States, 
\citet{Chiswick1995} also do this for Canada, Australia and Israel.
\citet{leslie+lindley2001} analyse both employment and income but
focus on a single year and English language only.
\citet{YaoandOurs2015} only analyse the role of Dutch language
on labour market performance in Netherlands.

The literature on Central Eastern Europe (CEE) and former Soviet Union
countries is rather limited.  We know that the different development
trajectories have lead to disparities both favoring Russian-speaking
minority in Ukraine \citep{Constant2011} and hindering them in Estonia
\citep{Leping2008}.  \citet{Kahanec2009} show that non-citizens suffer
less favorable outcomes, in particular ethnic Russians in Estonia and
Latvia. 
\cite{Alan2015} show that Russian language proficiency is valuable in
the former USSR republic in Caucasus (Armenia, Azerbaijan and
Georgia) by increasing employment likelihood between six and nine
percentage points.  \citet{Toomet2011}, however, fails to find any
income premium from the majority language fluency in Estonia and
Latvia.  This outcome is corroborated by \citet{Lindemann2013} who shows that Estonian fluency
is associated with faster job finding rate while it does not
explain the lower occupational prestige of Russian speakers.  
A number of these studies have to overcome severe data limitations,
such as lack of fine-grained information on ethnic background in
EU-wide surveys \citep{Kahanec2010}.

In this paper, we analyse the relationship between both
majority and minority language skills, and English for both
wage and unemployment.  We analyse Estonia, a former Soviet republic
and now a rapidly growing and
transforming Easter European country.  It is characterized by a stark linguistic division
where the Russian-speaking community forms almost 30\% of
the population.  We use Estonian Labor Force Survey, one of the
higher-quality long-term datasets, very well suitable for our
purpose.  We provide long--term trends from the last years
of the planned economy in 1989 till the post-recession time in 2012.

We find that the Estonian language, the majority and the official
language of the country, is associated with very
small albeit slowly increasing wage premium over
the years.  Interestingly, this is true only for women, Russian men enjoy
virtually no premium from the majority language skills.
Russian skills, on the other hand, are clearly related to
higher wage but the relationship is slowly fading.  English fluency was
associated with a very large wage premium during the early years of
economic transition, later its importance
declined but remained sizeable.  
In contrast, the sole language that plays a role in determining
unemployment probability is Estonian.  Its fluency is associated with
approximately 5 percentage points lower probability of being unemployed.
These results suggest that Estonian labour market is      
segmented both along gender and the along ethnic background, with men
more likely occupied in less communication-intense jobs.

The paper continues as follows: The next section provides a brief
background of Estonian society and labour market. In the following
section we discuss the data and provide some descriptive
statistics. In \cref{sec:method} we describe the econometric
approach, sections \ref{sec:results} and \ref{sec:extensions} present
the results for unemployment probability and wages.  The two last
sections, \ref{sec:segregation-analysis} and \ref{sec:discussion}, are
devoted to additional analysis of labor market segregation, and
concluding discussion.

@@

Such an influx of mainly
Russian-speaking population
led to two \emph{de facto} official languages
by 1970s.
The society was largely divided across language, and the division was
also reflected at workplace \citep{leppik+vihalemm2015JofBaltStud}.
Certain sectors in the economy, such as the armed forces, railways and the
merchant fleet
were dominated by
Russian-speaking workers while more local industries,
such as agriculture and
public administration,
were dominated by ethnic Estonians.
Both communities followed the media in their own respective language, the
universities taught many subjects in both Russian and Estonian,
and service sector was largely bilingual.  However, the bilingualism
was mainly one-sided as the bulk of Estonians spoke Russian, but not
the other way around.  A particular institution in parts of the Soviet
Union was segregated school system, where Russian schools existed along the
schools teaching in the local language.  Hence a common education
system, one of the central integration mechanisms in the developed world,
was (and still largely is) missing in Estonia.
The widening use of Russian also
caused increasing concerns about the future of the Estonian
language, in turn leading to  unwillingness to use Russian by
the ethnic Estonians.  This further strengthened the language-based segregation.

When the central Soviet authority weakened during the late 1980, the
Estonian-speaking community was quick to organize, and established a
nation-state in 1991 where
Estonian was granted status as the sole official language.  The
above-mentioned ``one-sided bilingualism'' started turning
around.  The universities quickly moved to Estonian-only curriculum,
Estonian language become a compulsory subject in Russian schools
while teaching Russian in Estonian-language schools declined.
These trends have continued till this day today, according to
\citet{HTM2015}, one quarter of the students from non
Estonian-speaking families studied in Estonian schools.
As a result, the younger generation of ethnic Estonians is
increasingly less fluent in Russian while the ethnic Russians are
steadily getting better in Estonian.

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s.  
\Cref{fig:gdp_growth} shows that grew averaged at approximately 7\%
annually
before the financial crisis in 2008, afterwards the figure has been around 3\%.
As a bi-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of transition.

\begin{figure}[h]
  \centering                      
  \caption{Annual GDP growth in Estonia}\label{fig:gdp_growth}
  \includegraphics[width=0.6\linewidth]{gdp_growth_per_year.eps}
  \caption*{Source: World Bank - World Development Indicators}
\end{figure}

@@

Our main analysis relies on the Estonian Labour Force Survey (ELFS),
first conducted in 1995 by the Estonian Statistical Office.  We limit
ourselves with years 2000--2012 in the main analysis, motivated by a
methodology break: Until 1999 the survey was conducted as an annual
cross-section and later as a quarterly rotating panel.  The survey
contains data on approximately 4000 distinct individuals annually,
however, due to it's rotating nature, we have roughly 16,000 annual
observations for each year since 2000.  For the long-term analysis we
also use the earlier waves.  These wave also include a retrospective
section that contains income up to several years back.

@@

The labor market outcomes we analyze are income, the ``last salary on the main job'' and
labor force status (working, unemployed or inactive).

The survey includes self-reported language skills on languages used at
home, and all other languages they understand.  The respondents are
given some guidance in assessing their skills, they can report either
no knowledge, understanding only, speaking but not writing, writing, or using the
language at home.  We consolidate this information into a dummy variable, equal
to one if one speaks or writes, or uses that language at
home, and zero otherwise.  Hence we disregard passive understanding
only.

For further evidence on ethnic and gender segregation at workplace we use data
from 2008 Integration Monitor Survey.  The survey was conducted by
SaarPoll and focuses on
ethnic relations in Estonia.  The sample size is 1500,
500 of the respondents consider themselves not ethnic Estonians.  We analyse
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''.  The first
question addresses the ethnic segregation while the second focuses on
the need of communication.

@@

The table indicates that there are substantial disparities between
Estonians and Russians in wage and unemployment. For instance, the
average monthly salary\footnote{When interpreting these figures,
  one should keep in mind that during this period, 2000-2012, the average monthly salary grow by 180\%,
  from \EUR{314} to \EUR{887} while the language knowledge changed rapidly as
  well.  However, the ethnic composition of the workforce remained
  roughly constant.} of Estonian
males who are not fluent in English is \EUR{532} while for Russians it
is only \EUR{452}. However, we also see that Russians who speak
English earn approximately 40\% more than those who do not (\EUR{452}
versus \EUR{326} monthly for men and \EUR{307} versus \EUR{221} for
women). Estonian skills are associated with lower income premium, 15\%
for men (\EUR{378} versus \EUR{329} monthly) and 30\% for women
(\EUR{260} versus \EUR{203}).  Analogously, English is more closely
associated with income for Estonians compared to Russian language. In
contrast, unemployment shows a different picture. For ethnic Russians,
it is Estonian language fluency that is more closely associated with
low unemployment, those fluent in Estonian have 11.2 percent
unemployment rate while for those not fluent it is 17.3 percent.  The
corresponding numbers for English are 11.2 and 15.1.  For Estonians,
English is more closely associated with unemployment than Russian:
Estonians fluent in English have in average 5.0 percent unemployment
while those who are not fluent have 8.8 percent.  The difference
related to Russian fluency is only 1.6 percentage points.  However,
these correlations are not easy to interpret as besides of the strong
time trends, English skills are
also related to education, metropolitan residence, and general
ability.

@@

The language fluency variables clearly show that ethnic identity is
closely related to language use: more than 99\% of Estonians report
being fluent in Estonian, the figure for Russians is similar. We
also see that Estonians are better in English than Russians and they
are also better in Russian than vice versa.

Average age reflects the trend of younger generations
being more fluent in English and Estonian while older Estonians being
better in Russian. The family status--related variables correspond to the
respective age distribution. Not surprisingly, inter-ethnic
households include disproportionally large share of Russian speaking
Estonians and Estonian speaking Russians.
The average experience follows the same pattern as age.
Public sector jobs are associated with disproportionally large share of
Estonian fluency among Russian workers, probably reflecting the formal
language requirements and little room for ethnic segregation in that
sector.

@@

We analyse the relationship between labour market outcomes and language
skills using the ordinary multivariate regression.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
We refrain from doing that. Our decision is mainly data driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age at immigration to instrument Estonian skills
	for those Russians who were in the country at 1990.  The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies.  For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only
report OLS results, and, even more, find that the available proxy
variables have little influence on the final results.

@@

We estimate three models that differ in terms of 
control variables. The first specification only includes the language skills
and year dummies.  It describes the ``raw'' effect of language skills,
net of strong time trends in our data.
The second specification additionally includes individual
characteristics (age, education) and family descriptors (marital
status, indicator for inter-ethnic marriage, having kids up to 17 years
old) and county of residence.  
Finally, the third specification also adds industry, occupation, and other
workplace descriptors.
Note that in case of unemployment probability,
we only estimate the first two models.

@@

As Estonian economy went through turbulent times and rapid changes
during the 1990s, it is instructive to analyze the effect long-term
trends in the estimated effect size.  Here we look at the yearly estimates from 1992 to 2012.
The reader should keep in mind that while the pre-2000 waves include retrospective income, they do not include retrospective language skills. In this way, the results before
1995 are partly based on extrapolation, assuming individual language
skills did not vary before that date.

We report the results for model 2, i.e. models that include year
dummies and individual controls, but not workplace
characteristics.\footnote{The results for models with workplace data (model
  3 for wage)
  look qualitatively similar.  More controls in the model reduces the effect size but does not qualitatively
  change it's temporal pattern.}

>>>>

\fxsetup{theme=color}
%Custom commands
\newcommand{\eur}{\EUR}
\newcommand{\std}[1]{\emph{(#1)}}
\newcommand{\V}{\ensuremath\checkmark}
\def\fns{\footnotesize}
\renewcommand*{\vec}[1]{\boldsymbol{#1}}

@@

%\maketitle
%\begin{center}
%	\textbf{{\large Language Skills in an Ethnically Segmented Labour market:
%				Estonia 1989 -- 2012}} \\
%	{\large Sven-Kristjan Bormann} \\
%	\emph{School of Economics and Business Administration, University of Tartu, Narva mnt. 4, 51007 Tartu, Estonia\\ Tel.: +372 5844 3873\\ Email: \href{mailto:bormann@ut.ee}{bormann@ut.ee}}	\\
%
%	{\large Svetlana Ridala} \\
%	\emph{School of Business and Governance, Tallinn University of Technology, Akadeemia tee 3, 12618 Tallinn, Estonia\\ Email: \href{
%			mailto:svetlana.ridala@gmail.com}{
%			svetlana.ridala@gmail.com}} \\
%		{\large Ott Toomet}\\
%		\emph{Information School, University of Washington, USA; Email: \href{mailto:otoomet@uw.edu}{otoomet@uw.edu}}
%%	\emph{(Received XX XXXX 20XX; accepted XX XXXX 20XX)}\\
%	\textbf{Corresponding author:} Sven-Kristjan Bormann Email: \href{mailto:bormann@ut.ee}{bormann@ut.ee}
%\end{center}
%\paragraph{Short biographical note:} \hspace{0pt} \\
%Sven-Kristjan Bormann born 28.11.1985, graduated in 2011 from Kiel University, Germany in Economics with specialisation in Quantitative Economics and Computer Science. Since September 2012 PhD-student in Economics at Tartu University, Estonia. \\
%Svetlana Ridala born 28.03.1976, graduated in 2003 from Tallinn Pedagogical University, Estonia master's (scien) degree in mathematics. Since 2009 PhD-student in Economics at Tallinn School of Economics and Business Administration, Estonia.
%\\
%Ott-Siim Toomet born 18.03.1970, graduated in 2004 from Aarhus University, Denmark PhD in Economics. Since 2015 Visiting scholar, Information School, University of Washington.
%
%\paragraph{Research funding}
%This research was supported by European Social Fund's Doctoral Studies
%and Internationalisation Program DoRa through Archimedes Foundation, and the Estonian Science Foundation grant 8627.
%\newpage
%\noindent
{\centering \textbf{Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012}}
\begin{abstract}
 \noindent
 \textbf{Purpose --} We analyse the relationship between skills of
 Estonian, Russian and English, and labour market
 outcomes in Estonia, a linguistically divided country.
 \\
 \textbf{Design/methodology/approach --} We use the Estonian Labour
 Force Surveys 1992--2012. We rely on multivariate linear regression
 models to document the relationship between language skills and
 labour market outcomes.
 \\
 \textbf{Findings --} Estonian language knowledge (for ethnic Russians)
 are important determinants of unemployment. Wage, in contrary, is
 closely related to English skills. Ethnic Russian men do not earn
 any premium from speaking Estonian, while women, fluent in Estonian
 earn approximately 10\% more. For ethnic Estonians, Russian fluency
 is associated with a similar income gain.
 \\
 \textbf{Research limitations/implications -- } Due to the
 observational nature of the data, the effects reported in our study
 are not causal effects. As a second limitation, the self-reported
 language skills data may be imprecise and hence the effects we
 report may be too small.
 \\
 \textbf{Practical implications --} Our results stress the role of
 workplace segregation, both along gender and ethnic lines, in
 determining the individual labour market experience. 
 \\
 \textbf{Originality/value --} We provide a comprehensive overview of
 the effects of language skills in a rapidly developing labour
 market in a linguistically divided economy. We analyse several
 languages with different legal status and document long-term
 trends in the effects.
 \\
\end{abstract}
\textbf{Keywords:} language skills, ethnicity, wage gap, unemployment, Estonia \\
\textbf{JEL classification:} J15, J31, J71

@@

There is also an increasing body of literature focusing on other languages than the local main business language. These studies generally find
that English is a significant predictor of better income
\citep{Lang2009, Casale2011, Toomet2011, Williams2011, azam+2013EDandCC, isphording2013, fabo+2017E}.
The role of other languages, in particular in multi-lingual economy, is
less clear. For instance,
\cite{Drinkwater1997} finds
a positive effect for Welsh speaking workers in
Wales and \cite{Armstrong2015} shows that bilingual native French-speakers enjoy
a significant premium both inside and outside of Quebec.
\citet{FrancoisGrin1998}, in contrast,
shows that Italian language skills in Swiss non-Italian speaking
regions are associated with a lower wage. It remains unclear if
this is a true effect or just an omitted variable bias.

Few previous studies have provided a comprehensive analysis of a labour
market from the viewpoint of language skills.
\citet{Chiswick1995, chiswick+miller2002, chiswick+miller2007,
 Bellante1998, Chiswick2010} analyse the effects of dominant language
fluency for earnings among immigrants in the United States, 
\citet{Chiswick1995} also do this for Canada, Australia and Israel.
\citet{leslie+lindley2001} analyse both employment and income but
focus on a single year and English language only.
\citet{YaoandOurs2015} only analyse the role of Dutch language
on labour market performance in the Netherlands.

The literature on Central Eastern Europe (CEE) and former Soviet Union
countries is rather limited. We know that the different development
trajectories have led to disparities both favouring Russian-speaking
minority in Ukraine \citep{Constant2011} and hindering them in Estonia
\citep{Leping2008}. \citet{Kahanec2009} show that non-citizens suffer
less favourable outcomes, in particular, ethnic Russians in Estonia and
Latvia. 
\cite{Alan2015} show that Russian language proficiency is valuable in
the former USSR republic in the Caucasus (Armenia, Azerbaijan and
Georgia) by increasing employment likelihood between six and nine
percentage points. \citet{Toomet2011}, however, fails to find any
income premium from the majority language fluency in Estonia and
Latvia. This outcome is corroborated by \citet{Lindemann2013} who shows that Estonian fluency
is associated with faster job finding rate while it does not
explain the lower occupational prestige of Russian speakers. 
A number of these studies have to overcome severe data limitations,
such as lack of fine-grained information on ethnic background in
EU-wide surveys \citep{Kahanec2010}.

In this paper, we analyse the relationship between both
majority and minority language skills, and English for both
wage and unemployment. We analyse Estonia, a former Soviet republic
and now a rapidly growing and
transforming Easter European country. It is characterized by a stark linguistic division
where the Russian-speaking community forms almost 30\% of
the population. We use Estonian Labour Force Survey, one of the
higher-quality long-term datasets, well suitable for our
purpose. We provide long--term trends from the last years
of the planned economy in 1989 till the post-recession time in 2012.

We find that the Estonian language, the majority and the official
language of the country, is associated with a very
small albeit slowly increasing wage premium over
the years. Interestingly, this is true only for women, Russian men enjoy
virtually no premium from the majority language skills.
Russian skills, on the other hand, are clearly related to
higher wage but the relationship is slowly fading. English fluency was
associated with a very large wage premium during the early years of
economic transition, later its importance
declined but remained sizeable. 
In contrast, the sole language that plays a role in determining
unemployment probability is Estonian. Its fluency is associated with 
approximately 5 percentage points lower probability of being unemployed.
These results suggest that the Estonian labour market is 
segmented both along gender and the along ethnic background, with men
more likely occupied in less communication-intense jobs.

The paper continues as follows: The next section provides a brief
background of Estonian society and the labour market. In the following
section, we discuss the data and provide some descriptive
statistics. In \cref{sec:method} we describe the econometric
approach, sections \ref{sec:results} and \ref{sec:extensions} present
the results for unemployment probability and wages. The two last
sections, \ref{sec:segregation-analysis} and \ref{sec:discussion}, are
devoted to an additional analysis of labour market segregation, and
concluding discussion.

@@

Such an influx of mainly
Russian-speaking population
led to two \emph{de facto} official languages
by 1970s.
The society was largely divided across language, and the division was
also reflected at workplace \citep{leppik+vihalemm2015JofBaltStud}.
Certain sectors in the economy, such as the armed forces, railways and the
merchant fleet
were dominated by
Russian-speaking workers while more local industries,
such as agriculture and
public administration,
were dominated by ethnic Estonians.
Both communities followed the media in their own respective language, the
universities taught many subjects in both Russian and Estonian,
and service sector was largely bilingual. However, the bilingualism
was mainly one-sided as the bulk of Estonians spoke Russian, but not
the other way around. A particular institution in parts of the Soviet
Union was a segregated school system, where Russian schools existed along the
schools teaching in the local language. Hence, a common education
system, one of the central integration mechanisms in the developed world,
was (and still largely is) missing in Estonia.
The widening use of Russian also
caused increasing concerns about the future of the Estonian
language, in turn leading to an unwillingness to use Russian by
the ethnic Estonians. This further strengthened the language-based segregation.

When the central Soviet authority weakened during the late 1980s, the
Estonian-speaking community was quick to organize and established a
nation-state in 1991 where
Estonian was granted status as the sole official language. The
above-mentioned ``one-sided bilingualism'' started turning
around. The universities quickly moved to Estonian-only curriculum,
the Estonian language became a compulsory subject in Russian schools
while teaching Russian in Estonian-language schools declined.
These trends have continued until today, according to
\citet{HTM2015}, one quarter of the students from non-Estonian-speaking families studied in Estonian schools.
As a result, the younger generation of ethnic Estonians is
increasingly less fluent in Russian while the ethnic Russians are
steadily getting better in Estonian.

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s. 
\Cref{fig:gdp_growth} shows that grew averaged at approximately 7\%
annually
before the financial crisis in 2008, afterwards the figure has been around 3\%.
As a by-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of transition.

\begin{figure}[h]
 \centering           
 \caption{Annual GDP growth in Estonia}\label{fig:gdp_growth}
 \includegraphics[width=0.6\linewidth]{Figure1.eps}
 \caption*{Source: World Bank - World Development Indicators}
\end{figure}

@@

Our main analysis relies on the Estonian Labour Force Survey (ELFS),
first conducted in 1995 by the Estonian Statistical Office. We limit
ourselves with years 2000--2012 in the main analysis, motivated by a
methodology break: Until 1999 the survey was conducted as an annual
cross-section and later as a quarterly rotating panel. The survey
contains data on approximately 4000 distinct individuals annually,
however, due to its rotating nature, we have roughly 16,000 annual
observations for each year since 2000. For the long-term analysis, we
also use the earlier waves. These waves also include a retrospective
section that contains income up to several years back.

@@

The labour market outcomes we analyse are income, the ``last salary on the main job'' and
labour force status (working, unemployed or inactive).

The survey includes self-reported language skills on languages used at
home and all other languages they understand. The respondents are
given some guidance in assessing their skills, they can report either
no knowledge, understanding only, speaking but not writing, writing, or using the
language at home. We consolidate this information into a dummy variable, equal
to one if one speaks or writes, or uses that language at
home, and zero otherwise. Hence, we disregard passive understanding
only.

For further evidence on ethnic and gender segregation at the workplace, we use data
from 2008 Integration Monitor Survey. The survey was conducted by
SaarPoll and focuses on
ethnic relations in Estonia. The sample size is 1500,
500 of the respondents consider themselves not ethnic Estonians. We analyse
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''. The first
question addresses the ethnic segregation while the second focuses on
the need for communication.

@@

The table indicates that there are substantial disparities between
Estonians and Russians in wage and unemployment. For instance, the
average monthly salary\footnote{When interpreting these figures,
 one should keep in mind that during this period, 2000-2012, the average monthly salary grow by 180\%,
 from \EUR{314} to \EUR{887} while the language knowledge changed rapidly as
well. However, the ethnic composition of the workforce remained roughly constant.} of Estonian
males who are not fluent in English is \EUR{532} while for Russians it
is only \EUR{452}. However, we also see that Russians who speak
English earn approximately 40\% more than those who do not (\EUR{452}
versus \EUR{326} monthly for men and \EUR{307} versus \EUR{221} for
women). Estonian skills are associated with lower income premium, 15\%
for men (\EUR{378} versus \EUR{329} monthly) and 30\% for women
(\EUR{260} versus \EUR{203}). Analogously, English language skills are more closely
associated with income for Estonians compared to Russian language skills. In
contrast, unemployment shows a different picture. For ethnic Russians,
it is Estonian language fluency that is more closely associated with
low unemployment, those fluent in Estonian have 11.2 percent
unemployment rate while for those not fluent it is 17.3 percent. The
corresponding numbers for English are 11.2 and 15.1. For Estonians,
English is more closely associated with unemployment than Russian:
Estonians fluent in English have on average 5.0 percent unemployment
while those who are not fluent have 8.8 percent. The difference
related to Russian fluency is only 1.6 percentage points. However,
these correlations are not easy to interpret as besides of the strong
time trends, English skills are
also related to education, metropolitan residence, and general
ability.

@@

The language fluency variables clearly show that ethnic identity is
closely related to language use: more than 99\% of Estonians report
being fluent in Estonian, the figure for Russians is similar. We
also see that Estonians are better at English than Russians and they
are also better at Russian than vice versa.

Average age reflects the trend of younger generations
being more fluent in English and Estonian while older Estonians being
better in Russian. The family status--related variables correspond to the
respective age distribution. Not surprisingly, inter-ethnic
households include a disproportionally large share of Russian speaking
Estonians and Estonian speaking Russians.
The average experience follows the same pattern as age.
Public sector jobs are associated with disproportionally large share of
Estonian fluency among Russian workers, probably reflecting the formal
language requirements and little room for ethnic segregation in that
sector.

@@

We analyse the relationship between labour market outcomes and language
skills using the ordinary multivariate regression.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
We refrain from doing that. Our decision is mainly data-driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age at immigration to instrument Estonian skills
	for those Russians who were in the country at 1990. The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies. For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only
report OLS results, and, even more, find that the available proxy
variables have little influence on the final results.

@@

We estimate three models that differ in terms of the control variables. The first specification only includes the language skills
and year dummies. It describes the ``raw'' effect of language skills,
net of strong time trends in our data.
The second specification additionally includes individual
characteristics (age, education) and family descriptors (marital
status, an indicator for inter-ethnic marriage, having kids up to 17 years
old) and county of residence. 
Finally, the third specification also adds industry, occupation, and other
workplace descriptors.
Note that in case of unemployment probability,
we only estimate the first two models.

@@

As the Estonian economy went through turbulent times and rapid changes
during the 1990s, it is instructive to analyse the effect long-term
trends in the estimated effect size. Here we look at the yearly estimates from 1992 to 2012.
The reader should keep in mind that while the pre-2000 waves include retrospective income, they do not include retrospective language skills. In this way, the results before
1995 are partly based on extrapolation, assuming an individual's language
skills did not vary before that date.

We report the results for model 2, i.e. models that include year
dummies and individual controls, but not workplace
characteristics.\footnote{The results for models with workplace data (model
 3 for wage)
 look qualitatively similar. More controls in the model reduce the effect size but does not qualitatively
 change its temporal pattern.}

The effect of language fluency on unemployment is shown on
Figure~\ref{fig:long-run_unemployment}, and on wage on
Figure~\ref{fig:long-run_wage}.
The results reveal a few interesting facts. First, the effects are
relatively stable over time, indicating that neither economic
environment nor the data quality have changed rapidly. Estonian
fluency has been associated with less unemployment for both Russian men and
women since around the year 2000, in the 1990s it had little effect. Russian fluency,
in contrast, had never had an impact on Estonians' unemployment probability.
